---
title: "MachineLearning"
author: "Petras Vaiciunas"
date: "January 31, 2016"
output: html_document
---
Practical Machine Learning Course Project
Raya Matoorian

November 20, 2015

Introduction
The goal of this project is to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. Here with prediction modeling tried to find patterns in their behavior and quantify how well they do the practices.


training data
test data
The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har.

```{r}
    setwd("C:/Users/pvaiciunas/Documents/R")
    training <- read.csv('./Data/pml-training.csv', na.strings=c("NA","#DIV/0!",""))
    testing  <- read.csv('./Data/pml-testing.csv',  na.strings=c("NA","#DIV/0!",""))
```

classe variable is the outcome and composed of five levels (A,B,C,D,E).

```{r}
    summary(training$classe)
    
##    A    B    C    D    E 
## 5580 3797 3422 3216 3607
```

Feature Selection
The dataset inclues variables that has less impact on prediction, thus we will exclude those variables with Near Zero Variance algorithm (NZV).

```{r}
    library(caret)
## Warning: package 'caret' was built under R version 3.2.2
## Loading required package: lattice
## Warning: package 'lattice' was built under R version 3.2.2
## Loading required package: ggplot2
## Warning: package 'ggplot2' was built under R version 3.2.2
# first we will remove variables with mostly NAs (threshold>75%)
    subTraining <- training
    for (i in 1:length(training)) {
      if (sum(is.na(training[ , i])) / nrow(training) >= 0.75) {
        for (j in 1:length(subTraining)) {
          if (length(grep(names(training[i]), names(subTraining)[j]))==1) {
            subTraining <- subTraining[ , -j]
          }
        }
      }
    }

# then we remove columns that are obviously not predictors
    subTraining2 <- subTraining[,8:length(subTraining)]

#remove variables with near zero variance
    NZV <- nearZeroVar(subTraining2, saveMetrics = TRUE)
    keep <- names(subTraining2)
    
```

Random Forest Model
Since Random Forest Model is appropriate for building classification problems, training data used to fit a model and then cross validated by test dataset.

```{r}
    library(randomForest)
## Warning: package 'randomForest' was built under R version 3.2.2
## randomForest 4.6-12
## Type rfNews() to see new features/changes/bug fixes.
# Random Forest Model
    set.seed(1234)
    modFit <- randomForest(classe~., data = subTraining2)
    print(modFit)
## 
## Call:
##  randomForest(formula = classe ~ ., data = subTraining2) 
##                Type of random forest: classification
##                      Number of trees: 500
## No. of variables tried at each split: 7
## 
##         OOB estimate of  error rate: 0.32%
## Confusion matrix:
##      A    B    C    D    E  class.error
## A 5576    4    0    0    0 0.0007168459
## B   13 3780    4    0    0 0.0044772189
## C    0   11 3410    1    0 0.0035067212
## D    0    0   21 3193    2 0.0071517413
## E    0    0    0    6 3601 0.0016634322
#cross validation with testing and training dataset
    predict_test  <- predict(modFit, testing,  type = "class")
    predict_train <- predict(modFit, training, type = "class")
    confusionMatrix(training$classe, predict_train)
## Confusion Matrix and Statistics
## 
##           Reference
## Prediction    A    B    C    D    E
##          A 5580    0    0    0    0
##          B    0 3797    0    0    0
##          C    0    0 3422    0    0
##          D    0    0    0 3216    0
##          E    0    0    0    0 3607
## 
## Overall Statistics
##                                      
##                Accuracy : 1          
##                  95% CI : (0.9998, 1)
##     No Information Rate : 0.2844     
##     P-Value [Acc > NIR] : < 2.2e-16  
##                                      
##                   Kappa : 1          
##  Mcnemar's Test P-Value : NA         
## 
## Statistics by Class:
## 
##                      Class: A Class: B Class: C Class: D Class: E
## Sensitivity            1.0000   1.0000   1.0000   1.0000   1.0000
## Specificity            1.0000   1.0000   1.0000   1.0000   1.0000
## Pos Pred Value         1.0000   1.0000   1.0000   1.0000   1.0000
## Neg Pred Value         1.0000   1.0000   1.0000   1.0000   1.0000
## Prevalence             0.2844   0.1935   0.1744   0.1639   0.1838
## Detection Rate         0.2844   0.1935   0.1744   0.1639   0.1838
## Detection Prevalence   0.2844   0.1935   0.1744   0.1639   0.1838
## Balanced Accuracy      1.0000   1.0000   1.0000   1.0000   1.0000
```
    
As the summary shows, the fitted model on trainig data has 100% accuracy. At the end, this model used to predict on test dataset and the output will write on a text file in the project directory.

```{r}
    predict_Final <- predict(modFit, testing, type = "class")
    print(predict_Final)
##  1  2  3  4  5  6  7  8  9 10 11 12 13 14 15 16 17 18 19 20 
##  B  A  B  A  A  E  D  B  A  A  B  C  B  A  E  E  A  B  B  B 
## Levels: A B C D E
    pml_write_files = function(x) {
        n = length(x)
        for (i in 1:n) {
            filename = paste0("problem_id_", i, ".txt")
            write.table(x[i], file=filename, quote=FALSE,
                        row.names=FALSE, col.names=FALSE)
        }
    }

    pml_write_files(predict_Final)
```